---
title: "Gov 50 Final Project: The Effects of Home Ice Advantage in the NHL"
author: "Thomas J. Mete"
description: "Exploring 'home advantage' in the NHL, this project investigates if playing at home increases goals scored, hypothesizing from the NHL's home ice advantage."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Section One: Introduction

In all major sports leagues, “home advantage” has emerged as an influential aspect regardless of the sport. It has left players, coaches, and insiders convinced the “home advantage” teams bear for fifty percent of their regular season games, may be enough to tilt the field in their favor when a game is underway. The notion of performing better at home is engrained in the culture of professional sports, serving as a coveted prize of teams that finish the season at the top of the standings when playoffs ensue. In a sports world that has become fiercely competitive on the field, off the field the rise of online gambling has altered the odds beating landscape. Viewers at home, turned recreational sports gamblers, like players are turning to “home advantage” as a potential prospect to help rise above the rest. Whether it be playing a game in front of a boisterous home crowd, the familiarity and comfort that comes with playing in a familiar stadium, or some intangible factor,  team’s have been left with the opinion that their local environment, due in fact, provide a crucial boost when it is all on the line.

Over the course of this project, I will empirically explore the effects of “home advantage” in the National Hockey League (NHL). I plan on exploring whether the location of the hockey game has an effect on the number of goals scored by a team with goals scored being a measurable and significant predictor of game outcomes.The central research question of this project is whether playing at home increases the number of goals scored by the team? I hypothesize that playing at home does indeed boost the goals scored of the home team, as unlike most other sports, in the NHL there is a built-in home advantage. This research question will be mainly tested and observed through the Boston Bruins.

During a stoppage in play, the final line change is given to the home team after the visiting team sends its players out onto the ice. This rule gives the coach of the home team more control and allows him to strategically deploy his players, either in a defensive posture or in an effort to create more offense. 

Exploring this research question — through the Boston Bruins — is of importance as it aims to empirically confirm or disprove a long standing belief with the NHL community that playing at home provides a boost. This understanding is crucial during playoff hockey games when teams are in a grueling battle to emerge as the league’s leading contenders and vie for the title of Stanley Cup Champions.



### Section Two: Data Section

#### Read in the Data Set 
```{r, eval = T}
library(dbplyr)
library(tidyverse)

# Read in the two data sets
latest <- read.csv("nhl_elo_latest.csv")
historical <- read.csv("nhl_elo.csv")

## Merge Datasets
total <- rbind(latest, historical)

## See the Datasets
knitr::kable(head(latest))
knitr::kable(head(historical))
```

##### Verify that the merge of datasets was successful.
```{r, eval = F}
total |>
  filter(season == 1918)

total |>
  filter(season == 2023)
```
##### Create a new variable to outline if a game concluded in overtime, multiple overtimes, or a shootout.
```{r}

## Make Overtime variable numeric
total <- total |>
  mutate(
    overtime = if_else(ot == "OT", 1, 0),
    shootout = if_else(ot == "SO", 1, 0)
  )
```

This project merges two data sets to analyze the effects of home ice advantage on goals scored over the entirety of the NHL. The first dataset contains historical NHL data beginning with the 1917-1918 NHL season and ending with the	2021-2022 NHL season. The second data set is the latest NHL data from the 2022-2023 NHL season. These datasets recorded every NHL game played across all seasons, the variables reported that are of interest to this project are home team name, away team name, home team score, away team score, whether the game was in the playoffs, if the game ended in overtime or a shootout. This data set contains data that is a cross-sectional research design. This is a cross-sectional research design as it collects data on the NHL games from a certain point in time and is strictly observational thus there are no interventions, manipulations or follow ups to the one time assessment.

In this research project I will be zooming in on one specific NHL team the Boston Bruins. I will be exploring the effects of home ice advantage with the following independent and dependent variables. The dependent variable (bos_score) in this study will be the number of goals scored by the Boston Bruins. This outcome variable will measure the effect of playing at home with more goals scored at home signifying a home ice advantage. The independent variable (bos_location) in this study is the location of the hockey game. This is the variable that will be manipulated (playing home or away) to observe the effect of the dependent variable which is the number of goals scored. These variables were measured through observations, the bos_location variable is either home or away, and the bos_score variable is a numeric variable that is a whole number greater than zero.

These two data sets did not provide team specific variables so I have created the independent and dependent variables for the Boston Bruins, along with other teams that will be discussed further in the project.

```{r}
bos_data <- total |>
  filter(home_team_abbr == "BOS" | away_team_abbr == "BOS") |>
  mutate(bos_location = if_else(home_team_abbr == "BOS", "Home", "Away"),
         bos_score = if_else(bos_location == "Away", away_team_score, home_team_score))
```

Here we have created a data set that has variables pertaining to the Boston Bruins. Below are the extracted columns of interest in the preliminary research of this project the independent and dependent variable. 

```{r}

bos_ind_dep <- bos_data[, c("bos_score", "bos_location")]
knitr::kable(head(bos_ind_dep))

```

A brief visualization of the independent and dependent variables can be found below as a boxplot. This provides us with a graphical representations of these two variables and provides a summary of the key characteristics of the data, allowing for easy identification of central tendency, dispersion, and the presence of outliers. A key reason I selected this method was it is useful to compare a catogrical (bos_location) and numerical (bos_score) variable. 

```{r}
bos_ind_dep_vis <- ggplot(bos_ind_dep, aes(x = bos_location, y = bos_score)) +
  geom_boxplot(fill = "#F5C710", color = "black", alpha = 0.7) +
  labs(title = "Boston Bruins: Boxplot Summary of Goals Scored by Game Location",
       x = "Game Location",
       y = "Goals Scored") +
  theme_minimal()
bos_ind_dep_vis
```


### Section Three: Results


XXXXXX

```{r}
#Find ATE of Playing at Home on Goals Scored
bos_ate <- bos_data |>
  group_by(bos_location) |>
  summarise(bos_score_mean = mean(bos_score)) |>
  pivot_wider(names_from = bos_location, values_from = bos_score_mean) |>
  mutate(ATE = `Home` - `Away`)
bos_ate

knitr::kable(
  bos_ate,
  col.names = c("Home", "Away", "ATE of Home Ice"), caption = "Boston Bruins: Number Goals Scored",
  digits = 2)

## Visualization Comparing Number of Goals Boston Scores at Home versus Away
plot_visb <- bos_data |>
  group_by(bos_location) |>
  summarise(bos_score_mean = mean(bos_score)) |>
  ggplot(aes(x = bos_location, y = bos_score_mean, fill = factor(bos_location))) +
  geom_col(width = 0.7) +
  labs(title = "Boston Bruins: Goals Scored Home vs Away",
       x = "Location",
       y = "Number of Goals Scored") +   
  scale_fill_manual(values = c("Home" = "#F5C710", "Away" = "black")) +
  geom_text(aes(label = round(bos_score_mean, 2)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, color = "black", size = 3)
plot_visb
```

##### Brief Interpretation of the Average Treatment Effect
The estimated effect of the Boston Bruins playing an NHL game at home versus playing an NHL game away is an increase in the Boston Bruins number of goals scored by 0.52.

```{r}
#Let's Try Bootstrapping
library(infer)

bos_boots <- bos_data |>
  rep_slice_sample(prop = 1, replace = TRUE, reps = 1000) |>
  group_by(replicate, bos_location) |>
  summarise(bos_score_mean = mean(bos_score)) |>
  pivot_wider(names_from = bos_location, values_from = bos_score_mean) |>
  mutate(ATE = `Home` - `Away`) |>
  select(replicate, ATE)

bos_ci_95 <- bos_boots |>
  get_confidence_interval(level = 0.95, type = "percentile")

bos_boots_plot <- bos_boots |>
  ggplot(mapping = aes(x = ATE)) + 
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.0055, fill = "#F5C710") +
  geom_vline(xintercept = unlist(bos_ci_95)) +
  labs(y = "Density", x = "Additional Goals Scored When Play at Home") + 
  ggtitle("Boston's Home Ice Advantage: The Effect on Goals over 1000 Bootstraps")
bos_boots_plot

knitr::kable(
  bos_ci_95,
  col.names = c("Lower Confidence Interval", "Upper Confidence Interval"), caption = "Boston Bruins: Confidence Interval at 95%",
  digits = 2)

```

##### Brief Interpretation of the Confidence Interval
The confidence interval for the Boston Bruins is 0.43 to 0.60, which does not include 0. This means that no difference between the number of goals scored at home or away is not a plausible value of the true population parameter, at least at the 95% confidence level. 

This confidence interval of 0.43 to 0.60 provides a range of plausible values in this case meaning the true difference in the number of goals scored at home versus away for the Boston Bruins. This also leads us to the conclusion that we have a statistically significant finding as a 95% confidence interval tells us that if we were to replicate this experiment, 95% of the results would exclude zero.


#### Looking at the Boston Bruins we will run a regression of the number of goals scored (`bos_score`) as the dependent variable, `bos_location` as the main independent variable. This is a two variable regression.

```{r}
library(broom)

##Regular Regression

fit_bos <- lm(bos_score ~ bos_location, data = bos_data)

summary(fit_bos)

var_labels_l <- c(
"(Intercept)" = "Intercept",
"bos_locationHome" = "Home Ice Advantage"
)

modelsummary::modelsummary(fit_bos,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels_l,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
                          
```

#### Looking at the Boston Bruins we will run a multi variable regression of the number of goals scored (`bos_score`) as the dependent variable, `bos_location` as the main independent variable and the following other control/independent variables playoff game (`playoff`) and game went to overtime (`ot`). 

```{r}
##Multi Variable Regression

fit_bos_multi <- lm(bos_score ~ bos_location + playoff + overtime + shootout, data = bos_data)

summary(fit_bos_multi)

var_labels <- c(
"(Intercept)" = "Intercept",
"bos_locationHome" = "Home Ice Advantage",
"playoff" = "Playoff Game",
"overtime" = "Game ends in Overtime",
"shootout" = "Game ends in Shootout"
)

modelsummary::modelsummary(fit_bos_multi,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

```

#### How does this change over fanbases that are considered less engaged vs fan bases that are hardcore? We wil use the Montrèal Canadiens as a hardcore fanbase and the Florida Panthers as a less engaged fan base, these choices were made off attendance.

```{r}
#What about in the best fan base (Montréal Canadiens)
#Filter to Only Montréal, Make Variable for if MTL Home or Away and Variable for MTL Goals
mtl_data <- total |>
  filter(home_team_abbr == "MTL" | away_team_abbr == "MTL") |>
  mutate(mtl_location = if_else(home_team_abbr == "MTL", "Home", "Away"),
         mtl_score = if_else(mtl_location == "Away", away_team_score, home_team_score))

#Find ATE of Playing at Home on Goals Scored
mtl_ate <- mtl_data |>
  group_by(mtl_location) |>
  summarise(mtl_score_mean = mean(mtl_score)) |>
  pivot_wider(names_from = mtl_location, values_from = mtl_score_mean) |>
  mutate(ATE = `Home` - `Away`)
mtl_ate
  
## Visualization Comparing Number of Goals Montrèal Scores at Home versus Away
plot_vism <- mtl_data |>
  group_by(mtl_location) |>
  summarise(mtl_score_mean = mean(mtl_score)) |>
  ggplot(aes(x = mtl_location, y = mtl_score_mean, fill = factor(mtl_location))) +
  geom_col(width = 0.7) +
  labs(title = "Montrèal Canadiens",
       x = "Location",
       y = "Number of Goals Scored") +   
  scale_fill_manual(values = c("Home" = "#DF5365", "Away" = "blue")) +
  geom_text(aes(label = round(mtl_score_mean, 2)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, color = "black", size = 3)

#What about in the worst fan base (Florida Panthers)
#Filter to Only Florida, Make Variable for if FLA Home or Away and Variable for FLA Goals
fla_data <- total |>
  filter(home_team_abbr == "FLA" | away_team_abbr == "FLA") |>
  mutate(fla_location = if_else(home_team_abbr == "FLA", "Home", "Away"),
         fla_score = if_else(fla_location == "Away", away_team_score, home_team_score))

#Find ATE of Playing at Home on Goals Scored
fla_ate <- fla_data |>
  group_by(fla_location) |>
  summarise(fla_score_mean = mean(fla_score)) |>
  pivot_wider(names_from = fla_location, values_from = fla_score_mean) |>
  mutate(ATE = `Home` - `Away`)
fla_ate
  
## Visualization Comparing Number of Goals Florida Scores at Home versus Away
plot_visf <- fla_data |>
  group_by(fla_location) |>
  summarise(fla_score_mean = mean(fla_score)) |>
  ggplot(aes(x = fla_location, y = fla_score_mean, fill = factor(fla_location))) +
  geom_col(width = 0.7) +
  labs(title = "Florida Panthers",
       x = "Location",
       y = "Number of Goals Scored") +   
  scale_fill_manual(values = c("Home" = "#F0E68C", "Away" = "grey")) +
  geom_text(aes(label = round(fla_score_mean, 2)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, color = "black", size = 3)

library(patchwork)

combined_plot1 <- plot_vism + plot_visf +
  plot_layout(ncol = 2) +
  plot_annotation(title = "Comparison of Goals Scored Home vs Away",
                  theme = theme(plot.title = element_text(hjust = 0.5)))
combined_plot1

# Combine the ATE table
library(kableExtra)
combined_ate <- cbind(fla_ate, mtl_ate) |>
  kbl(col.names = c("Home (FLA)", "Away (FLA)", "ATE of Home Ice (FLA)",
                    "Home (MTL)", "Away (MTL)", "ATE of Home Ice (MTL)"),
      caption = "Florida Panthers and Montréal Canadiens: Comparison of Number of Goals Scored",
      digits = 2) |>
  kable_styling()
combined_ate
```

##### Brief Interpretation of the Average Treatment Effect
The estimated effect of the Montrèal Canadiens playing an NHL game at home versus playing an NHL game away is an increase in the Montrèal Canadiens number of goals scored by 0.68.

The estimated effect of the Florida Panthers playing an NHL game at home versus playing an NHL game away is an increase in the Florida Panthers number of goals scored by 0.30.

```{r}
#Let's Try Bootstrapping
library(infer)

mtl_boots <- mtl_data |>
  rep_slice_sample(prop = 1, replace = TRUE, reps = 1000) |>
  group_by(replicate, mtl_location) |>
  summarise(mtl_score_mean = mean(mtl_score)) |>
  pivot_wider(names_from = mtl_location, values_from = mtl_score_mean) |>
  mutate(ATE = `Home` - `Away`) |>
  select(replicate, ATE)

mtl_ci_95 <- mtl_boots |>
  get_confidence_interval(level = 0.95, type = "percentile")

mtl_boots_plot <- mtl_boots |>
  ggplot(mapping = aes(x = ATE)) + 
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.0055, fill = "#DF5365") +
  geom_vline(xintercept = unlist(mtl_ci_95)) +
  labs(y = "Density", x = "Additional Goals Scored When Play at Home") + 
  ggtitle("Montréal Canadiens")

#Let's Try Bootstrapping
library(infer)

fla_boots <- fla_data |>
  rep_slice_sample(prop = 1, replace = TRUE, reps = 1000) |>
  group_by(replicate, fla_location) |>
  summarise(fla_score_mean = mean(fla_score)) |>
  pivot_wider(names_from = fla_location, values_from = fla_score_mean) |>
  mutate(ATE = `Home` - `Away`) |>
  select(replicate, ATE)

fla_ci_95 <- fla_boots |>
  get_confidence_interval(level = 0.95, type = "percentile")

fla_boots_plot <- fla_boots |>
  ggplot(mapping = aes(x = ATE)) + 
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.0055, fill = "#F0E68C") +
  geom_vline(xintercept = unlist(fla_ci_95)) +
  labs(y = "Density", x = "Additional Goals Scored When Play at Home") + 
  ggtitle("Florida Panthers")

# Combine the CI table
combined_plot2 <- mtl_boots_plot + fla_boots_plot / plot_layout(ncol = 2) + plot_annotation(
  title = "Home Ice Advantage: The Effect on Goals over 1000 Bootstraps",
  theme = theme(plot.title = element_text(hjust = 0.5))
)
combined_plot2

combined_ci <- cbind(fla_ci_95, mtl_ci_95) |>
  kbl(col.names = c("Lower Confidence Interval (FLA)", "Upper Confidence Interval (FLA)",
                    "Lower Confidence Interval (MTL)", "Upper Confidence Interval (MTL)"),
      caption = "Florida Panthers and Montréal Canadiens: Confidence Interval at 95%",
      digits = 2) |>
  kable_styling()
combined_ci

```

##### Brief Interpretation of the Confidence Intervals
The confidence interval for the Montrèal Canadiens is 0.59 to 0.76, which does not include 0. This means that no difference between the number of goals scored at home or away is not a plausible value of the true population parameter, at least at the 95% confidence level.

The confidence interval for the Florida Panthers is 0.18 to 0.42, which does not include 0. This means that no difference between the number of goals scored at home or away is not a plausible value of the true population parameter, at least at the 95% confidence level.

These confidence intervals provide a range of plausible values in this case meaning the true difference in the number of goals scored at home versus away for the specified team. This also leads us to the conclusion that we have a statistically significant finding as a 95% confidence interval tells us that if we were to replicate this experiment, 95% of the results would exclude zero.

















